---
title: "A practical guide to spatial interaction modelling"
format:
  REGION-pdf:
    keep-tex: true
    pdf-engine: pdflatex
    docstatus: final
  REGION-html:
    toc: false
    format-links: false
author:
  - name: Blinded
    affiliations:
      - name: 
        city: 
        country: 
abstract: |
  This document is only a demo explaining how to use the template.
keywords: [spatial interaction modelling, gravity modelling, flow data]
bibliography: bibliography.bib  
#ojsnum: 555
#jvol: 11
#jnum: 1
#jpages: 100--200
#jyear: 2024
#jauthor: blinded
#received: January 1, 2024
#accepted: January 2, 2024
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| echo: false
# clear the environment
rm = list(ls())
```

# Introduction {#sec-intro}

Spatial interaction models (SIMs) is a core tool to simulate flows
between different locations in physical space. They are a valuable
resource through which the geographic structure between locations
encoded in aggregate flows of people, information and goods can be
represented and understood. Intuitively, SIMs seek to capture the
spatial interaction between places as a function of three components:
origin attributes, destination attributes and their separation. Inspired
by Newtonian concepts developed in physics, spatial flows between
locations are conceived as the result of their proportional
gravitational force and inverse association with spatial separation.
Attributes of origin and destination locations are employed to represent
gravitational forces pushing and pulling people, information and goods
between specific locations. Various forms of distance and costs are used
to represent the deterring effects of geographical separation on spatial
flows.

SIMs are widely used for prediction and inference. They are used to make
inference about the factors contributing to influence spatial flows.
They have been used to understand the magnitude and direction of
influence of individual and place-level attributes on geographic flows.
Understanding the effect of these factors offers valuable evidence to
inform the development of appropriate plans, strategies and
interventions [@fotheringham_okelly1989]. SIMs are also used to make
predictions of the size of spatial flows. These predictions are normally
used to assess the impact of interventions and creation of "what-if"
scenarios, providing guidance for the identification of optimal
locations and size for potential new service units \[REF\]. In this
context, SIMs are often used to evaluate the impact of new bus stops,
shopping stores, schools or housing units on their potential demand and
traffic changes [@fotheringham_okelly1989]. To these ends, SIMs have
been used to address questions in a variety of settings, including
retail, migration, transport, trade, commuting, school travel and more
broadly urban planning.

Yet, the implementation of SIMs remains a challenge. Algorithms to
calibrate the parameters of SIMs have remained locked away, either
behind dense algebraic notation in dusty papers from the 1970s, or
behind paywalls of commercial software [@rowe2024]. Additionally,
@rowe2024 noted a dearth of knowledge within geographical education as
SIMs are not widely taught in undergraduate programmes in the same way
as, for instance, regression models are taught in economics or social
psychology. This situation is argued to have occurred despite the
availability of effective routines to calibrate SIMs via popular linear
and general linear modelling frameworks, and as practical expediency is
sacrificed at the expense of theoretical or technical prowess
[@rowe2024]. The ways in which calibration procedures are presented as
lengthy mathematical derivation or passing reference to ordinary least
square tend to hamper accessibility for the easy implementation of SIMs.

This computational notebook contributes to redressing these issues. It
aims to provide an intuitive, understandable and practical guide to
estimate SIMs in a variety of modelling frameworks. It will include the
necessary code to calibrate SIMs, using origin-destination
travel-to-work data for the United Kingdom in R programming language.
The code provided is generisable and can be adapted to different
origin-destination flow data and contexts, including migration, student,
transport, trade, currency, data transfer, vessel, shipment and freight
flows.

The notebook is structured as follows. The next section sets out some
fundamental concepts and definitions relating to SIMs. @sec-comenv
identifies the libraries used before @sec-data describes the data.
@sec-visualising illustrates key techniques to visualise complex spatial
interaction data, and @sec-modelling shows and explains how to estimate
SIMs using a range of modelling frameworks. It start with traditional
mathematical and Ordinary Least Squares (OLS) approaches to more
advanced statistical frameworks, such as Generalised Linear Mixed Models
(GLMMs) and machine learning algorithms.

# Context

SIMs take various forms. Newtonian gravity models are probably the most
widely known and used form of SIMs. Inspired by Newton’s law of gravity,
the basic gravity version of these models assumes that the spatial flows
or interactions between an origin ($i$) and a destination $j$ is
proportional to their masses ($M_{i}$ and $M_{j}$) and inversely
proportional to their separation ($D_{ij}$). Locations are expected to
interact in a positively reinforcing manner that is multiplicative of
their masses, but to diminish with the intervening role of their
separation. The separation between locations is often represented by a
distance decay function and is measured in terms of the distance, cost
or time involved in the interaction. Generally, the model includes a
constant ($G$) ensuring that the expected flows do not exceed their
respective observed counts, and a parameter ($k$) representing the
deterring effect of geographical separation. The task is to estimate
these parameters. Formally a gravity model can be expressed as:

$$
T_{ij} = G \frac{M_{i} M_{j}}{D^{k}_{ij} }
$$

SIMs have three key inputs: (1) a matrix of flows between a set of
origins and destinations; (2) a measure of separation between origins
and destinations; and, (3) measures of masses at origin and destination
locations. The literature usually considers a family of SIMs taking four
forms which refers to various constraints placed on parameters of the
model [@wilson1971]. There is an *unconstrained* version which is
actually constrained to ensure that the total sum of the predicted flows
from a gravity model be equal the total sum of the observed flows across
all origins and destinations. Constrained versions are used to ensure
that specific origin or destination observations are met. Three general
formulations of constrained models are used: *production-constrained*,
*attraction-constrained* and *doubly-constrained* models.
*Production-constrained* versions are used to constrain a model so that
the predicted flows emanating from individual origins is equal to their
respective observed numbers. *Attraction-constrained* versions do the
same but at individual destinations. *Doubly-constrained* versions
combine these two sets of constraints to ensure predicted flows are
equal to observed flows at both origins and destinations.

# Computation environment {#sec-comenv}

```{r}
#| warning: false
library(tidyverse)
library(sf)
library(glmmTMB)
```

# Data {#sec-data}

Commuting or migration?

# Visualising spatial interaction data {#sec-visualising}

# Estimating spatial interaction models {#sec-modelling}

## Mathemathical gravity models

## Statistical gravity models

## Extensions

### Generalised linear mixed gravity models

### Machine learning gravity models

# References
